---
title: "index"
author: "ADG"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
---

# Intro

## Getting Started

Data pulled from Morningstar Direct on Jan. 24, 2023. Files combined using _combine_files.r_. 
Data in M* filtered:

* != Fund of funds: Yes
* != US Category: Money Market

57,451 rows, 117 columns

Added column _Obsolete_ to indicate if share class is in existence or obsolete. I assume a blank _Obsolete..Date_ means the share class still exists.

Also added _Inception_Year_ and _Obsolete_Year_ because these variables are used below.

Data filtered to remove:

* "Load Waived" Share.Class.Type
* Share classes launched and liquidated during the same 18 month period
* Share classes launched after 2022

48,761 rows, 121 columns

```{r setup, message = FALSE, warning = FALSE}

library(tidyverse) #For Data Analysis
library(lubridate) #For working with dates
library(DT) #For visualizing tables


Full <- read_csv("data_combined.csv",
                 guess_max = 20000) %>%
  rename_all(make.names) %>%
  mutate(
    Inception_Year = year(Inception..Date),
    Obsolete_Year = year(Obsolete..Date),
    Obsolete =
      case_when(!is.na(Obsolete..Date) ~ "Obsolete",
                TRUE ~ "Exists"),
    Life_Span = time_length(interval(Inception..Date, Obsolete..Date), "month")
  ) %>%
  filter(
    Share.Class.Type != "Load Waived" |
      is.na(Share.Class.Type),
    Life_Span >= 18 |
      is.na(Obsolete_Year) |
      is.na(Inception_Year),
    Inception_Year < 2023)

```

# Overview

```{r graph_overview, echo=FALSE}

ggplot(data = Full, mapping = aes(x = year(Inception..Date), fill = Obsolete)) +
  geom_bar(position = "stack") +
  ggtitle("Share Class Launches") +
  ylab("Count") +
  xlab("YEAR") +
  theme_classic() +
  theme(plot.title = element_text(size = 11, face = "bold"))

```

## Net Count

There was a large drop off in launches in 2020 compared to recent years. However it appears rationalization continued at roughly the same levels as recent years.


```{r data_overview, echo = TRUE, message = FALSE}

#A function to count the number of share classes either created or liquidated each year

Year_Count <- function(colz){
 Full %>%
    group_by_at(colz) %>% 
    summarise(Count = n(), .groups = "drop") %>% 
    arrange(desc(.[[1]]))
    }
    

inception_year <- Year_Count(colz = c("Inception_Year")) #Tallys share classes by year created
obsolete_year <- Year_Count(colz = c("Obsolete_Year")) #Tallys share classes by year liquidated

merge_type <- Year_Count(colz = c("Obsolete_Year", "Obsolete..Type")) #Tallys share classes by year liquidated and liquidation type


#Counts the net number of share classes created/cut

Net_Count <- full_join(inception_year, obsolete_year, by = c("Inception_Year" = "Obsolete_Year"),
                       suffix = c("_inception", "_obsolete")) %>% 
  rename(Year = Inception_Year) %>% 
  group_by(Year) %>% 
  mutate(Net_Count = sum(Count_inception, -Count_obsolete, na.rm = TRUE))
```

```{r data_overview_tbl, echo = FALSE}

datatable(Net_Count, colnames = c('Share_Classes_Launched' = 3, 'Share_Classes_Cut' = 4))

```

## New Funds

The data from Morningstar is at the share class level. This is a method to look at launches at the fund level.

```{r new_funds, message=FALSE}

#This looks at when a fund's oldest share class was created
#This assumes that the oldest share class's inception date is equal to the fund's creation date

New_Funds <- Full %>% 
  filter(Oldest..Share.Class == "Yes") %>% 
  group_by(Inception_Year, Index..Fund) %>% 
  summarise(Count = n(), .groups = "drop") %>% 
  mutate(
    Pct_Change = round(
      ((Count/lag(Count) - 1) * 100),
      1
    )) %>% 
  arrange(desc(Inception_Year))

```


```{r new_funds_graphs, echo=FALSE}

ggplot(data = New_Funds, mapping = aes(x =Inception_Year, y=Count, color=Index..Fund)) +
  geom_line() +
  ggtitle("New Fund Launches") +
  ylab("Count") +
  xlab("YEAR") +
  theme_classic() +
  theme(plot.title = element_text(size = 11, face = "bold"))

New_Funds %>% 
  select(-Pct_Change) %>% 
  pivot_wider(names_from = Index..Fund, values_from = Count) %>%
  rename(
    Passive = Yes,
    Active = No
    ) %>% 
  group_by(Inception_Year) %>% 
  mutate(total = sum (Active, Passive, na.rm = TRUE)) %>% 
    datatable()

```

## New Funds (Alternative)

The method above uses the inception date of the _Oldest..Share.Class_ as a proxy for when a fund was created. This method uses _FundId_.

The oldest inception date related to a _FundID_ is a proxy for when the fund was created. Or the oldest obsolete date related to a a _FundID_ (with no active share classes) is likely when that fund was liquidated/merged.

There were some cases identified in the data where a FundID had no corresponding _Oldest..Share.Class_. This could be an error or signal that a fund's oldest share class is not captured in the data from Morningstar.

The results for 2020 ended up equal using the method above.

This method shows there were 15,825 funds in 2020 vs 15,677 in 2019 -- a change of 148 funds.
This method shows there were 15,464 in 2018 -- a change of 213 funds.


```{r fund_id, echo = FALSE, message = FALSE}

# a function to count share classes launched/liquidated before a certain date, grouped by their FundId

status_count <- function(date_v) {
  #prefix for column names created by pivot_wider
  prefix <- paste("Count_", year(date_v), "_", sep = "")
  #variable to later call Aldready_Obsolete column
  obnow <- paste("Count_", year(date_v), "_", "Already_Obsolete", sep = "")
  #variable to later call Becomes_Obsolete column
  oblater <- paste("Count_", year(date_v), "_", "Becomes_Obsolete", sep = "")
  #variable to later call Not_Obsolete column
  exsts <- paste("Count_", year(date_v), "_", "Not_Obsolete", sep = "")
  #dataframe
  Full %>%
    #filter to look at only funds that were created before a certain date
    filter((Inception..Date < date_v | is.na(Inception..Date))) %>%
    #creates new column to determine if a share class becomes obsolete after date_v
    mutate(
      Obsolete_Past =
        case_when(
          #obsolete dates on or after date_v mean share class becomes obsolete
          Obsolete..Date >= date_v ~ "Becomes_Obsolete",
          #obsolete dates before date_v mean share class already obsolete
          Obsolete..Date < date_v ~ "Already_Obsolete",
          #blank obsolete dats mean the fund is not obsolete
          is.na(Obsolete..Date) ~ "Not_Obsolete",
          #if above criteria not met throws "ERROR"
          TRUE ~ "ERROR"
        )
    ) %>%
    group_by(FundId, Obsolete_Past) %>%
    #counts how many share classes in each fund (FundID) are obsolete using Obsolete_Past
    summarise(Count = n(), .groups = "drop") %>%
    #Turns dataframe wider so each Obsolete_Past catagory has its own column
    pivot_wider(
      names_from = Obsolete_Past,
      values_from = Count,
      names_prefix = c(prefix)
    )  %>%
    #Adds New Column that determines the overall status of the fund based on share classes
    mutate(
      !!paste0("Fund_Status_", year(date_v)) :=
        case_when(
          #Zero Not_Obsolete or Becomes_Obsolete share classes with at least one Already_Obsolete sc equals an Obsolete_Fund
          is.na(!!sym(exsts)) &
            is.na(!!sym(oblater)) &
            !!sym(obnow) >= 1 ~ "Obsolete_Fund",
          #Zero Not_Obsolete share classes with at least 1 Becomes_Obsolete sc equals a fund that Becomes_Obsolete
          is.na(!!sym(exsts)) &
            !!sym(oblater) >= 1 ~ "Obsolete_Later",
          #At least one share class that is Not_Obsolete means the fund exists as of date_v
          !is.na(!!sym(exsts)) ~ "Exists_Fund",
          TRUE ~ "Other"
        ))
  }


#a breakdown of active vs. obsolete funds as of Dec. 31, 2022
status_2022 <- status_count("2023-01-01")
#a breakdown of active vs. obsolete funds as of Dec. 31, 2021
status_2021 <- status_count("2022-01-01")
#a breakdown of active vs. obsolete funds as of Dec. 31, 2020
status_2020 <- status_count("2021-01-01")

#a breakdown of active funds as of date
status_summary <- function(x){
  x %>% 
    group_by_at(5) %>% 
    count(name = "Count")
  }


#a breakdown of active funds as of Dec. 31, 2022
status_summary_2022 <- status_summary(status_2022)

#a breakdown of active vs. obsolete funds as of Dec. 31, 2021
status_summary_2021 <- status_summary(status_2021)

#a breakdown of active vs. obsolete funds as of Dec. 31, 2020
status_summary_2020 <- status_summary(status_2020)

```
# Companies

To look at which firms cut or added share classes I first clean up the company names with Branding.Names.Mod. 

```{r companies, echo = TRUE}


company_change <- Full %>%
  mutate(
    Branding.Name.Mod =
      case_when(
        grepl("State Street", Branding.Name) ~ "State Street",
        grepl("TIAA", Branding.Name) ~ "TIAA/Nuveen",
        grepl("Nuveen", Branding.Name) ~ "Morgan Stanley",
        grepl("Eaton Vance", Branding.Name) ~ "Morgan Stanley",
        grepl("Calvert", Branding.Name) ~ "Eaton Vance/Calvert",
        grepl("iShares", Branding.Name) ~ "iShares/BlackRock",
        grepl("BlackRock", Branding.Name) ~ "iShares/BlackRock",
        grepl("PowerShares", Branding.Name) ~ "PowerShares/Invesco",
        grepl("Invesco", Branding.Name) ~ "PowerShares/Invesco",
        grepl("DWS$", Branding.Name) ~ "DWS/Xtrackers",
        grepl("Xtrackers", Branding.Name) ~ "DWS/Xtrackers",
        grepl("^Capital Group", Branding.Name) ~ "American Funds",
        Branding.Name %in% c("Harding Loevner", "Third Avenue",
                             "Tweedy, Browner") ~ "AMG",
        Branding.Name == "Oppenheimer" ~ "PowerShares/Invesco",
        Branding.Name == "Wells Fargo" ~ "Allspring",
        Branding.Name %in% "Amundi" ~ "Pioneer Investments",
        TRUE ~ Branding.Name
      ))
    
    

    
    company_level_2017 <- company_change %>%
      filter(Inception_Year <= 2017) %>% 
      mutate(
        Count_2017 =
          case_when(
            is.na(Obsolete_Year) ~ "Exists_2017",
            Obsolete_Year <= 2017 ~ "Obsolete_2017",
            Obsolete_Year > 2017 ~ "Exists_2017",
            TRUE ~ "Other"
      ))
    
    company_level_2017 <- company_level_2017 %>%
      select(Branding.Name.Mod,
                Name,
                FundId,
                Count_2017) %>%
      group_by(Branding.Name.Mod, Name, Count_2017) %>%
      summarise(Tally_2017 = n()) %>%
      pivot_wider(names_from = Count_2017,
                  values_from = Tally_2017)
    

      

      company_level_2022 <- company_change %>%
      filter(Inception_Year <= 2022) %>% 
      mutate(
        Count_2022 =
          case_when(
            is.na(Obsolete_Year) ~ "Exists_2022",
            Obsolete_Year <= 2022 ~ "Obsolete_2022",
            Obsolete_Year > 2022 ~ "Exists_2022",
            TRUE ~ "Other"
      ))
    
    company_level_2022 <- company_level_2022 %>%
      select(Branding.Name.Mod,
                Name,
                FundId,
                Count_2022) %>%
      group_by(Branding.Name.Mod, Name, Count_2022) %>%
      summarise(Tally_2022 = n()) %>%
      pivot_wider(names_from = Count_2022,
                  values_from = Tally_2022)
    
    company_level_change <-
      full_join(company_level_2017, company_level_2022)
    
    company_level_change <- company_level_change %>%
      group_by(Branding.Name.Mod) %>%
      summarise(Exists_2022 = sum(Exists_2022, na.rm = T),
                Exists_2017 = sum(Exists_2017, na.rm = T),
                Change = sum(Exists_2022,-Exists_2017, na.rm = T))
    
    
    
    company_funds <- company_change %>%
      group_by(FundId) %>%
      mutate(
        Newest_ObDate = max(Obsolete_Year),
        Oldest_InDate = min(Inception_Year)
      ) %>%
      group_by(Branding.Name.Mod,
               FundId,
               Newest_ObDate,
               Oldest_InDate) %>%
      summarise() %>%
      mutate(
        Count_2022 =
          case_when(
            is.na(Newest_ObDate) ~ 1,
            Newest_ObDate == 2023 ~ 1,
            Newest_ObDate <= 2022 ~ 0
          ),
        Count_2017 =
          case_when(
            is.na(Newest_ObDate) & Oldest_InDate <= 2017 ~ 1,
            Oldest_InDate > 2017 | Newest_ObDate <= 2017 ~ 0,
            Newest_ObDate > 2017 ~ 1
          )
      )
    
        
    company_fund_change <- company_funds %>%
      group_by(Branding.Name.Mod) %>%
      summarise(
        Total_2017 = sum(Count_2017, na.rm = T),
        Total_2022 = sum(Count_2022, na.rm = T),
        Difference = sum(Total_2022,-Total_2017, na.rm = T)
      )
    
    company_count <- function(x, colz) {
      company_change %>%
        group_by_at(colz) %>%
        summarise(Count = n()) %>%
        ungroup() %>%
        arrange(desc(.[[2]]))
    }
    
    comp_inception <-
      company_count(colz = c("Branding.Name.Mod", "Inception_Year"))
    comp_obsolete <-
      company_count(colz = c("Branding.Name.Mod", "Obsolete_Year"))
    
    comp_new_funds <-
      company_count(
        colz = c("Branding.Name.Mod", "Inception_Year", "Oldest..Share.Class")
      ) %>%
      filter(Oldest..Share.Class == "Yes")
    
    comp_old_funds <-
      company_count(
        colz = c("Branding.Name.Mod", "Obsolete_Year", "Oldest..Share.Class")
      ) %>%
      filter(Oldest..Share.Class == "Yes")
    
    comp_rank <- function(x) {
      x %>%  filter(.[[2]] == "2022") %>%
        mutate(Rank = rank(-Count)) %>%
        filter(Rank <= 20) %>%
        arrange(Rank)
    }
    
    top_cutters <- comp_rank(comp_obsolete)
    top_old_cutters <-  comp_rank(comp_old_funds)
    
    top_launchers <- comp_rank(comp_inception)
    top_new_launchers <- comp_rank(comp_new_funds)
    
```


```{r level, echo=FALSE}

datatable(company_level_change)

```

## Top Launchers

(Share Class)

```{r launchers, echo=FALSE}

datatable(top_launchers)

```

## Top New Fund Launchers

(Oldest Share Class)

```{r new_launchers, echo=FALSE}

datatable(top_new_launchers)

```


## Top Cutters

(Share Class)

```{r cutters, echo=FALSE}

datatable(top_cutters)

```

## Top Fund Cutters

(Oldest Share Class)

```{r old_cutters, echo=FALSE}

datatable(top_old_cutters)

```
# Share Class Overview

##Launches

```{r share_class_overview, message=FALSE}

simp_count <- function(x, colz){
  Full %>%  
    group_by_at(colz) %>% 
    count(name = "Count")
  }
```

From Morningstar Direct

```{r share_class_overview_graphic, echo=FALSE}

share_class_add_count <- simp_count(colz = c("Inception_Year", "Share.Class.Type"))

ggplot(data = share_class_add_count, aes(x = Inception_Year, y = Count, fill = Share.Class.Type)) +
  geom_bar(stat = "identity") + 
  scale_x_continuous(breaks = seq.int(from = 1900, to = 2021, by = 50)) +
  scale_y_continuous(breaks = seq.int(from = 0, to = 500, by = 250)) +
  facet_wrap(~Share.Class.Type) +
  ggtitle("Launches (Share Class Type)") +
  ylab("Count") +
  xlab("YEAR") +
  theme_classic() +
  theme(plot.title = element_text(size = 11, face = "bold"))
```

## 2022 Launches

```{r share_class_2022_launches, echo=FALSE}

share_class_add_count  %>% 
  filter(Inception_Year == "2022") %>% 
  arrange(desc(Count)) %>% 
  datatable()

```

##Liquidations

```{r share_class__cuts_overview_graphic, echo=FALSE}

share_class_cut_count <- simp_count(colz = c("Obsolete_Year", "Share.Class.Type"))

ggplot(data = share_class_cut_count, aes(x = Obsolete_Year, y = Count, fill = Share.Class.Type)) +
  geom_bar(stat = "identity") + 
  scale_x_continuous(breaks = seq.int(from = 1900, to = 2020, by = 50)) +
  scale_y_continuous(breaks = seq.int(from = 0, to = 1000, by = 250)) +
  facet_wrap(~Share.Class.Type) +
  ggtitle("Cuts (Share Class Type)") +
  ylab("Count") +
  xlab("YEAR") +
  theme_classic() +
  theme(plot.title = element_text(size = 11, face = "bold"))

```

## 2022 Liquidations

```{r share_class_2022_cuts, echo=FALSE}

share_class_cut_count  %>% 
  filter(Obsolete_Year == "2022") %>% 
  arrange(desc(Count)) %>% 
  datatable()

```
## Other/NA

This is an attempt to determine the share class type in cases where Morningstar has the share classs listed as other or NA. 

```{r share_clas_view, echo=FALSE}

#The data from Morningstar is dirty. This pattern pulls the last letters/digits in the mutual fund name to identify the share class.

PATTERN <-
  c(
    "(\\s)?([:alnum:]+)?([:alnum:][:punct:][:alnum:] LW)?( Load Waived)?([:alpha:]+ [:digit:]+)?( LW)?([:punct:]+)?([:alnum:]+)?(-[:alnum:])?(™)?(\\{[:alnum:]+\\})?([:punct:]..[:punct:])?(\\s)?([:punct:][:alnum:]+[:punct:])?$"
  )

#The data from Morningstar is dirty. This pattern refines the pattern above to identify the share class.

CUT <-
  c(
    "[:alpha:]+ (?=529.+)|([\\{\\(][:alnum:]+[\\}\\)])|([:punct:]\\X\\X$)|(-)|(\\sLW)|(\\sLoad\\sWaived)|(\\([:alpha:]+\\))"
  )

#The data from Morningstar is dirty. This pattern makes manual changes to share classification for when the patterns above failed.

CHANGE <- c(
  "^Adm(in$)?(in[:alpha:]+)?$" = "Admin",
  "[:alnum:]+R6$" = "R6",
  "(?!AA)[:alpha:]+A$" = "A",
  "A\\{m" = "A",
  "([:alpha:]+)([:punct:])?([:alpha:]+)?A$" = "A",
  "([:alpha:]+)([:punct:])?([:alpha:]+)?C$" = "C",
  "PlusC([:digit:]+)$" = "C",
  "([:alpha:]+)?([:Punct:][:alpha:]+)?Inst(l$)?(itutional$)?$" = "Instl",
  "([:alnum:]+)?Ins$" = "Instl",
  "([:alnum:]+)?Insl$" = "Instl",
  "[:alnum:]+Ist$" = "Instl",
  "I{1}" = "Instl",
  "([:alpha:]+)?([:punct:])?([:alpha:]+)?Y$" = "Y",
  "([:alpha:]+)?([:punct:])?([:alpha:]+)?R$(?<!(SDR$)|(IR$))" = "R",
  "[:alnum:]+R5$" = "R5",
  "([:alpha:]+)?([:Punct:][:alpha:]+)?In([vest]+)?(ment$)?$(?<!(Int$))" = "Investor",
  "([:alpha:]+)?([:punct:])?([:alpha:]+)?N$" = "N",
  "([:alpha:]+)?([:punct:])?([:alpha:]+)?T$" = "T",
  "[:alnum:]+R4$" = "R4",
  "([:alpha:]+)?([:Punct:][:alpha:]+)?Adv(iser$)?(isory$)?$" = "Advisor",
  "([:alpha:]+)?([:Punct:][:alpha:]+)?ADV$" = "Advisor",
  "([:alpha:]+)?([:punct:])?([:alpha:]+)?K$" = "K",
  "([:alpha:]+)?([:punct:])?([:alpha:]+)?F$" = "F",
  "[:alnum:]+R3$" = "R3",
  "[:alnum:]+F3$" = "F3",
  "([:alpha:]+)?([:punct:])?([:alpha:]+)?IS$" = "IS",
  "Retire(ment)?$" = "R",
  "(?![FIVS]+I$|([:alpha:]+)?II$)([:alpha:]+)?I$(?<!(IR$))" = "I",
  "([:alpha:]+)([:punct:])?([:alpha:]+)?I$(?<!(F/I$))" = "I"
)


#Using the above patterns we can clean the share class information
#Name -> Extract -> New -> Revision

INC_Data <- Full %>%
  select(Name) %>%
  mutate(
    Match = str_detect(Name, PATTERN),
    EXTRACT = str_extract(Name, PATTERN),
    New = str_trim(str_replace_all(EXTRACT,  CUT, "")),
    Revision = str_replace_all(New, CHANGE)
    )


#This counts the number of share class types in the data
INC_Summary <- INC_Data %>%
  group_by(Revision) %>%
  summarise(cases = n()) %>%
  arrange(-cases)


#This  takes the Full dataframe and adds new columns showinbg how we detected the share class (EXTRACT, New, Revision)
#In cases where Share.Class.Type is either NA or Other we subsistute Revision
INC_MODIFIED <- Full %>%
  mutate(
    Match = str_detect(Name, PATTERN),
    EXTRACT = str_extract(Name, PATTERN),
    New = str_trim(str_replace_all(EXTRACT,  CUT, "")),
    Revision = str_replace_all(New, CHANGE),
    Share_Class_Mod =
      case_when(
        is.na(Share.Class.Type) ~ Revision,
        Share.Class.Type == "Other" ~ Revision,
        Share.Class.Type == "Retirement" ~ Revision,
        TRUE ~ Share.Class.Type
      ))



```

## Changes

```{r share_class_overview_changes, echo=FALSE}

INC_Data %>% 
  select(-Match, everything()) %>% 
  datatable()

```

## Inception Year
(2022)

```{r share_class_inception_dt, echo = FALSE}

INC_MODIFIED %>%
  select(Name, Share.Class.Type, Revision, Share_Class_Mod, Inception_Year, Obsolete_Year) %>% 
  filter(Inception_Year == 2022 | Inception_Year == 2021) %>% 
  group_by(Share_Class_Mod, Inception_Year) %>%
  summarise(cases = n()) %>%
  arrange(-cases)  %>% 
  pivot_wider(values_from = cases, names_from = Inception_Year) %>% 
  view()

```

## Obsolete Year

(2022)

```{r share_class_obsolete_dt, echo = FALSE}

INC_MODIFIED %>%
  select(Name, Share.Class.Type, Revision, Share_Class_Mod, Inception_Year, Obsolete_Year) %>% 
  filter(Obsolete_Year == 2022 | Obsolete_Year == 2021) %>% 
  group_by(Share_Class_Mod, Obsolete_Year) %>%
  summarise(cases = n()) %>%
  arrange(-cases)  %>% 
  pivot_wider(values_from = cases, names_from = Obsolete_Year) %>% 
  view()

```

